Final Project - Intro to R - Jeffrey Yim

I will be using my own data that I obtained in a retrospective study of patients undergoing echocardiography (cardiac ultrasound) and right heart cath (invasive measurements of pressures in your pulmonary arteries, blood vessels supplying the lung.) My goal was to see whether non-invasive measurements of using cardiac ultrasound (RVFWSL) is correlated with invasive measurements using right heart cath (mean pulmonary pressures)



```{r}
#loading the necessary libraries
library(tidyverse)
library(haven)
library(here)
library(janitor)
library(gtsummary)
```

```{r}
data <- read_dta(here::here("data", "strain.dta")) #reading in my data as 'data' variable using here::here, which construct file paths with reference to the top directory holding my .Rproj file.

```

```{r}
data <- data %>% #I am taking my data object and using pipe operator to overwrite the data with the result
  mutate(rhcheight_meter = as.numeric(rhcheight) / 100) %>% #create a new variable of height in meters
  mutate(bmi = (as.numeric(rhcweight)/rhcheight_meter^2)) #create a new variable of BMI

bmi_summary <- data %>% #creating a new variable to summarize bmi
  group_by(sex) %>% #separating the bmi by sex
  summarize(avg_bmi = mean(bmi, na.rm = TRUE)) #getting the mean BMI while ignoring missing value 

bmi_summary 
```

```{r}
data <- data %>% 
  filter(!is.na(rhcheight)) %>%  #Filter out those with missing height
  filter(!is.na(rhcweight)) %>% #Filter out those with missing weight
  mutate(bsa=(sqrt(as.numeric(rhcheight)*as.numeric(rhcweight)/3600))) #create a new variable of BSA. used as.numeric as my data was from stata and kept using another label

summary(data$age)

data <- data %>% 
  mutate(age_fct = case_when( #Creating age_fct variable to categorize into factors
    age <35 ~ 1,
    age >= 35 & age < 50 ~ 2,
    age >= 50 & age < 70 ~ 3,
    age >= 70 ~ 4
    ))

data <- data %>% 
  mutate(
    age_cat = factor(
      age_fct, 
      levels = c(1, 2, 3, 4), #assigning levels
      labels = c("Young", "Young-middle", "Old-middle", "Old") #labels to each factor
    ),
    age_cat = fct_relevel(age_cat, "Young", "Young-middle", "Old-middle", "Old"),#relevel to order from youngest -> oldest
    age_cat = fct_recode( #recode labels
      age_cat, 
      "<35 (years)" = "Young",
      "35-49 (years)" = "Young-middle",
      "50-69 (years)" = "Old-middle",
      ">70 (years)" = "Old"
    )
  )


summary(data$age_cat) #taking a look at the summary

data_clean <- data %>% 
  filter(
    is.finite(rvfwsl),          # Keeps only finite values, i got a warning saying there were 'non-finite' values outside scale range when trying to run on data
    is.finite(preavrtmpap))

bsa_summary <- data %>% 
  summarize(avg_bsa = mean(bsa, na.rm = TRUE), .by = age_cat ) #summarizing the BSA by age category, ignoring missing

bsa_summary

data |> 
  tabyl(sex, age_cat) |> #making a table using tabyl function with sex and age_cat varialbes
  adorn_percentages() |> #add on percentages
  adorn_pct_formatting() #change formatting to make it %

table_data <- data %>% #storing new variable to create data on select variables
  filter(age >=18, !is.na(bmi) & bmi >0, !is.na(preavrtmpap)) %>%  # Age criteria & valid BMI values
  select(sex, age, age_cat, bmi, bsa, ends_with("SL"), mpap20, preavrtmpap) #using select and select helper
```

Going to create the table 1

```{r}
table1 <- table_data %>% 
  tbl_summary(
    by = sex, #stratify by sex
    label = list(
      age_cat ~ "Age Category", #re-label the categories
      bmi ~ "Body Mass Index",
      bsa ~ "Body Surface Area",
      rvfwsl ~ "RV Free Wall Strain",
      rv4csl ~ "RV Four Chamber Strain",
      mpap20 ~ "Pulmonary Hypertension",
      preavrtmpap ~ "Mean PA Pressures (mmHg)"
    )
  ) %>% 
  add_p(test = list( #adding p-values 
    all_continuous() ~ "t.test", 
    all_categorical() ~ "fisher.test")) %>% 
  add_overall() %>%  #add overall numbers
  bold_labels() %>%  #bold the labels
  modify_caption("**Patient Characteristics**") #add a title


table1 #print the table
gtsummary::as_gt(table1) %>% gt::gtsave(here("figures", "table 1.pdf")) #save the table as a pdf
```

Creating regression models

```{r}
linear_model <- lm(preavrtmpap ~ rvfwsl, data = data_clean) #creating linear regression model using mpap and rvfwsl
summary(linear_model) #print summary

library(broom)
tidy(linear_model) #helps tidy models
```

Creating 3 figures

```{r}
# Plot 1: Boxplot of Pulmonary Pressures by Sex
ggplot(data_clean, aes(x = sex, y = preavrtmpap, fill = sex)) +  # Set up plot with x as sex, y as pa pressures (preavrtmpap, fill by sex
  geom_boxplot() +  # Add boxplots
  labs(
    title = "Distribution of Pulmonary Pressures by Sex",  # Set plot title
    x = "Sex",  # Label for x-axis
    y = "Pulmonary Pressures (mmHg)"  # Label for y-axis
  ) +
  theme_minimal() +  # Use a clean, minimal theme
  theme(
    legend.position = "none", # Remove the legend (sex is already on x-axis)
    plot.title = element_text(hjust = 0.5, face = "bold"))  # Center and bold title


ggsave(here("figures", "Figure 1.pdf")) #saving it as Figure 1
```

```{r}
# Plot 2: Histogram of PH
ggplot(data_clean, aes(x = age_cat, fill = factor(mpap20))) + #create ggplot using factor of the mpap20 variable as the fill
  geom_bar(position = "stack") +  # Shows counts 
  facet_grid(cols = vars(sex)) +  # Splits by sex
  labs(
    title = "Distribution of Pulmonary Hypertension by Age and Sex", # Set plot title
    x = "Age Category", # Set x-axis title
    y = "Number of Pulmonary Hypertension",  # Set y-axis title
    fill = "Pulmonary Hypertension"  # fill by PH status
  ) +
  scale_fill_discrete( labels = c("No PH", "PH")) +  # rename 0&1 to "No PH"/"PH"
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold")  # Center and bold title
  )

ggsave(here("figures", "Figure 2.pdf"))
```

```{r}
# Plot 3: Scatter plot between RVFWSL and MPAP
library(ggpmisc)

regression<- ggplot(data_clean, aes(x = rvfwsl, y = preavrtmpap, color = sex)) + #specifying which variables i want in plot
  geom_point() +  # Scatter plot
  geom_smooth(method = "lm", se = TRUE) +  # Adds regression line + 95% CI
  labs(
    x = "Right Ventricular Free Wall Strain", #re-labelling
    y = "Mean PA Pressure (mmHg)",
    color = "Sex",
    title = "RV Free Wall Strain vs Mean PA Pressures"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))  # Center and bold title)
    
regression

ggsave(here("figures", "Figure 3.pdf"))
```







